Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Fix case in example
Typos

It turns out however, that in presence of higher-order contracts union
and intersection is deceptively difficult. The literature on
higher-order contracts with union and intersection, while keenly aware
of the fact, doesn't really explain why. We point and illustrate the
problems and trade-offs inherent to union and intersection contracts,
via example and a survey of the literature.
\end{abstract}

@@

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\typescript{null} (\python{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,KeilThiemannUnionIntersection}.

@@

Let us pause for a moment and consider the following: it is Nickel's
ambition to be able to manipulate configurations like Nixpkgs. With
over 50\,000 packages, is one of the largest repository of software
packages in existence~\cite{repology}. Concretely, Nixpkgs is a
dictionary mapping packages to build recipes. That is, a massive,
over-50\,000-key-value-pair wide dictionary. It is absolutely out of the
question to evaluate the entirety of this dictionary every time one
needs to install 10 new packages: this would result in a painfully
slow experience.

@@

    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let mkHost = fun server ext => [(*@{\ldots}@*)] in
  |              -----------------------
  |              evaluated to this
  = ++, 2nd argument
\end{lstlisting}
    \caption{Error reporting without contract}
    \label{fig:contract-reporting-wo}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
let catHosts | Str -> Str = fun last =>
  [(*@{\ldots}@*)]
\end{lstlisting}
\vspace{10ex}
    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | --------------------------- (2) calling <func>
    \end{lstlisting}
    \caption{Error reporting with contract}
    \label{fig:contract-reporting-w}
  \end{subfigure}
  \caption{Contracts improve error messages}
\end{figure*}

@@

This same property is desirable of contracts as well, otherwise we
lose an important benefit of dynamic checking. Preferably, we want to
be able to extend the universe of contracts with user-defined
predicates (sometimes called flat contracts).

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal types
\footnote{Note that in TypeScript, union and intersection are written as
\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
as we do in Nickel.}.
What value could we fill the square ($\square$) with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{42 | "Nickel"} or a
\typescript{"Nickel" | 3.5}. In general,
when using a union, we have to treat it as if it were any one
of the alternatives. In the case of the example, we should provide
it with a value that could be pushed to both arrays, the only possible one
is \typescript{"Nickel"},
since it's the only value that satisfies the intersection of
\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.

@@

Yet intersections are useful in their own right. They can be used to
combine dictionaries in the style of object-oriented inheritance. For instance, in
Figure \ref{fig:intersection-record} two types are defined
\nickel{Animal} and \nickel{Pet}, and a variable that is compatible
with both types is declared, with type \nickel{Animal @& Pet}. This
particular application is supported, for instance, by TypeScript.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let c = contracts.frompred (fun f =>
  f 0 == 0) in
let g | (Str -> Str) @& c
      = fun x => x
\end{lstlisting}
\caption{intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}

>>>>

It turns out however, that in presence of higher-order contracts union and
intersection are deceptively difficult. The literature on higher-order contracts
with union and intersection, while keenly aware of the fact, doesn't really
explain why. We point and illustrate the problems and trade-offs inherent to
union and intersection contracts, via example and a survey of the literature.
\end{abstract}

@@

Union types, meaning a type \nickel{A @| B} containing values which
belong either to a type \nickel{A} or \nickel{B}, are a popular tool
when adding static types to a dynamic language. In particular, both
TypeScript~\cite{TypeScriptUnions} and MyPy~\cite{MyPyOptional}, use
union types to model the frequent practice to use the value
\typescript{null} (\python{None} in Python) to represent an absent
optional value. This is why the gradual typing literature, concerned
with formalising the interplay between static and dynamic type
systems, has been quite interested in union
types~\cite{RootCauseOfBlame,gradualCastagna,ORTIN2011278,ToroTanterGradualUnion,KeilThiemannUnionIntersection}.

@@

Let us pause for a moment and consider the following: it is Nickel's
ambition to be able to manipulate configurations like Nixpkgs. With
over 50\,000 packages, it is one of the largest repository of software
packages in existence~\cite{repology}. Concretely, Nixpkgs is a
dictionary mapping packages to build recipes. That is, a massive,
over-50\,000-key-value-pair wide dictionary. It is absolutely out of the
question to evaluate the entirety of this dictionary every time one
needs to install 10 new packages: this would result in a painfully
slow experience.

@@

    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let makeHost = fun server ext => [(*@{\ldots}@*)] in
  |                --------------------
  |                evaluated to this
  = ++, 2nd argument
\end{lstlisting}
    \caption{Error reporting without contract}
    \label{fig:contract-reporting-wo}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
let catHosts | Str -> Str = fun last =>
  [(*@{\ldots}@*)]
\end{lstlisting}
\vspace{10ex}
    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | --------------------------- (2) calling <func>
    \end{lstlisting}
    \caption{Error reporting with contract}
    \label{fig:contract-reporting-w}
  \end{subfigure}
  \caption{Contracts improve error messages}
\end{figure*}

@@

This same property is desirable of contracts as well, otherwise we
lose an important benefit of dynamic checking. Preferably, we want to
be able to extend the universe of contracts with user-defined
predicates.

@@

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}
We have shown how values of a union are built, is worth mentioning
how they are used.
For instance, consider the TypeScript function on Figure \ref{fig:addElem},
we want it to insert an element to a list that is either an
\typescript{Array< 42 | "Nickel" >} or an \typescript{Array< "Nickel" | 3.5 >},
where \typescript{42}, \typescript{3.5}, and \typescript{"Nickel"} are literal
types\footnote{Note that in TypeScript, union and intersection are written as
\typescript{|} and \typescript{&}, instead of \nickel{@|} and \nickel{@&},
as we do in Nickel.}.
What value could we fill the square ($\square$) with for this to typecheck?
Since we don't know what's the shape of \typescript{list},
we can't know whether to push a \typescript{42 | "Nickel"} or a
\typescript{"Nickel" | 3.5}. In general,
when using a union, we have to treat it as if it were any one
of the alternatives. In the case of the example, we should provide
it with a value that could be pushed to both arrays, the only possible one
is \typescript{"Nickel"},
since it's the only value that satisfies the intersection of
\typescript{42 | "Nickel"} and \typescript{"Nickel" | 3.5}.

@@

Yet, intersections are useful in their own right. They can be used to
combine dictionaries in the style of object-oriented inheritance. For instance, in
Figure \ref{fig:intersection-record} two types are defined
\nickel{Animal} and \nickel{Pet}, and a variable that is compatible
with both types is declared, with type \nickel{Animal @& Pet}. This
particular application is supported, for instance, by TypeScript.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let c = contracts.fromPred (fun f =>
  f 0 == 0) in
let g | (Str -> Str) @& c
      = fun x => x
in g 0
\end{lstlisting}
\caption{intersection and user defined contracts}
\label{fig:inter-flat-contracts}
\end{figure}
